---
title: "PH525.5x - Intro to Bioconductor - installs"
author: "Michael Hunt"
date: "April 5, 2016"
output: html_document
---

To get the most out of this course, you should be very familiar with R and Rstudio, and should be able to install and load packages as they are mentioned.  To verify this, try the following in your R session.

```{r}
library(Biobase)
sessionInfo()
```

The R version should be 3.2.2 or 3.2.3.  If the library() command fails, you should perform a basic Bioconductor installation. Do the following:

```{r}
source("http://www.bioconductor.org/biocLite.R")
biocLite()
```

A collection of core packages will be installed.

A package developed specifically for this course should be obtained using

```{r}
library(devtools)
install_github("genomicsclass/ph525x")
```
u
In order to carry out the tasks of the first week, you will need to have some fairly large packages installed, including these Bioconductor packages:

`IRanges`
`TxDb.Hsapiens.UCSC.hg19.knownGene`
`GenomicFeatures`

`BSgenome.Hsapiens.UCSC.hg19`
`SNPlocs.Hsapiens.dbSNP.20120608`
`Homo.sapiens`

Several of these packages are quite large, for example BSgenome.Hsapiens is 656MB and SNPlocs.Hsapiens is 211 MB.  If you have a slow internet connection you will want to install these well in advance.

As you work through exercises, if a Bioconductor package not found error is encountered, use

```{r}
library(BiocInstaller)
biocLite("Homo.sapiens")
```