---
title: "PH525 4x HDDA: Smoothing Exercises"
author: "Michael Hunt"
date: "April 2, 2016"
output: html_document
---

### Smoothing Exercises #1

Use the data generated in question 2.6.2 (Conditional Expectations)

```{r}
n = 10000
set.seed(1)
men = rnorm(n,176,7) #height in centimeters
women = rnorm(n,162,7) #height in centimeters
y = c(rep(0,n),rep(1,n))
x = round(c(men,women))
##mix it up
ind = sample(seq(along=y))
y = y[ind]
x = x[ind]
```

Set the seed at 5, `set.seed(5)` and take a random sample of 250 individuals from the population like this:

```{r}
set.seed(5)
N = 250
ind = sample(length(y),N)
Y = y[ind]
X = x[ind]
```
Use `loess` to estimate $f(x)=E\left (Y|X=x\right )$ using the default parameters. What is the predicted $f(168)$?

```{r}
fit <- loess(Y~X, degree=1, span=1/3)

newx <- seq(min(X),max(X),len=100) 
smooth <- predict(fit,newdata=data.frame(X=newx))
library(rafalib)
mypar ()
plot(X,Y,col="darkgrey",pch=16)
lines(newx,smooth,col="black",lwd=3)

smooth[newx==168]
```
__Model answer:__

```{r}
fit=loess(Y~X)
predict(fit,newdata=data.frame(X=168))

##Here is a plot
xs = seq(160,178)
Pr =sapply(xs,function(x0) mean(Y[X==x0]))
plot(xs,Pr)
fitted=predict(fit,newdata=data.frame(X=xs))
lines(xs,fitted)
```

### Smoothing Exercises #2

The `loess` estimate above is a random variable thus we should compute its standard error. Use Monte Carlo simulation to compute the standard error of your estimate of $f(168)$.

Set the seed to 5, `set.seed(5)` and perform 1000 simulation of the computations performed in question 2.7.1. Report the the SE of the `loess` based estimate.

```{r}
Nsim<-1000
N = 250
set.seed(5)
preds<-replicate(Nsim,{
ind = sample(length(y),N)
Y = y[ind]
X = x[ind]
fit=loess(Y~X)
predict(fit,newdata=data.frame(X=168))
})
hist(preds) # take a look
library(rafalib)
popsd(preds)
```

Model answer:

```{r}
##plot plots are optional
library(openintro)
data(COL)
set.seed(5)
B = 1000
N = 250
xs = seq(160,178)
plot(xs,xs,ylim=c(0,1),type="l") # empty chart
res = replicate(B,{
  ind = sample(length(y),N)
  Y = y[ind]
  X = x[ind]
  fit=loess(Y~X)
  ##optional plots
  fitted=predict(fit,newdata=data.frame(X=xs))
  lines(xs,fitted,col='#569BBD')
  estimate = predict(fit,newdata=data.frame(X=168))
  return(estimate)
  })
library(rafalib)
popsd(res)
```