---
title: "Untitled"
author: "Michael Hunt"
date: "April 20, 2016"
output: html_document
---

## Understanding and building R packages

[Book chapter: Understanding and building R packages](http://genomicsclass.github.io/book/pages/bioc2_rpacks.html#skel)

### What is an R package?
Conceptually, an R package is a collection of functions, data objects, and documentation that coherently support a family of related data analysis operations.

Concretely, an R package is a structured collection of folders, organized and populated according to the rules of Writing R Extensions.


### A new software package with package.skeleton

```{r juxta}
juxta = function (chrname="chr22", ...) 
{
    require(ERBS)
    data(HepG2)
    data(GM12878)
    require(ggbio)
    require(GenomicRanges)  # "subset" is overused, need import detail
    ap1 = autoplot(GenomicRanges::subset(HepG2, seqnames==chrname))
    ap2 = autoplot(GenomicRanges::subset(GM12878, seqnames==chrname))
    tracks(HepG2 = ap1, Bcell = ap2, ...)
# alternative code for Gviz below
#    require(Gviz)
#    ap1 = AnnotationTrack(GenomicRanges::subset(HepG2, seqnames==chrname))
#    names(ap1) = "HepG2"
#    ap2 = AnnotationTrack(GenomicRanges::subset(GM12878, seqnames==chrname))
#    names(ap2) = "B-cell"
#    ax = GenomeAxisTrack()
#    plotTracks(list(ax, ap1, ap2))
}
```
Basic idea of Juxta is to juxtapose pictures of the estrogen-like receptor alpha binding peaks for the two cell lines that we have isolated.

```{r}
library(ERBS)
juxta("chr22", main="ESRRA binding peaks on chr22")
```
or with a different chromosome...

```{r}
library(ERBS)
juxta("chr2", main="ESRRA binding peaks on chr2")
```

Now use __package.skeleton__ to make a package with this. The aim is Visualization of binding patterns where we juxtapose the results for two different cell lines.

So how do we make a package that manages the software for this visualization? Simple!

```{r}
?package.skeleton
```
we need to name the package then define the R packages that are going to be in there.

```{r}
package.skeleton("erbsViz",list="juxta")
install.packages("erbsViz", repos=NULL, type="source")
```
Use the erbsViz package as created during the video to perform the following computation:

```{r}
library(erbsViz)
jdemo = juxta()
```

What is the class of jdemo?

Answer:
```{r}
class(jdemo)
```

### Customization elements for the Tracks display

Have a look at getSlots(getClass(class(jdemo))) to get a sense of how the display might be customized.

What is the name of the Tracks slot that alters the color of the label text?

```{r}
getSlots(getClass("Tracks"))
```

### setup for a makeOrganismPackage call

In the preceding video, we saw how to make an `OrganismDb` package for S. cerevisiae. In this problem set, you will make such a package for C. elegans. The `makeOrganismPackage` function in `OrganismDbi` requires a list for the `graphData` parameter

```{r}

# gd = list( join1 = c(GO.db="GOID", org.Sc.sgd.db="GO"),
#            join2 = c(org.Sc.sgd.db="ENTREZID",
#               TxDb.Scerevisiae.UCSC.sacCer3.sgdGene="GENEID"))

gd = list(join1 = c(GO.db="GOID", org.Ce.eg.db="GO"),join2 = c(org.Ce.eg.db="ENTREZID", TxDb.Celegans.UCSC.ce6.ensGene="GENEID"))
```

What should you substitute for [fixme] in the call above? Hint: the package name to be used has eg.db as suffix. See the annotation [cheat sheet](http://genomicsclass.github.io/book/pages/annoCheat.html) for the contraction pattern.

Answer:  
`org.Ce.eg.db`

### A query to the new Cen.ele6 package

After correctly specifying gd above, use

```{r}
library(OrganismDbi)
makeOrganismPackage("Cen.ele6", gd, "C. elegans", "1.0.0","me <me@abc.com>", "me <me@abc.com>", ".")
```
and install the package via

```{r}
install.packages("Cen.ele6", 
    repos=NULL, type="source")
```
use library(Cen.ele6)

```{r}
library(Cen.ele6)
```
What is sum(seqlengths(Cen.ele6))?:

```{r}
sum(seqlengths(Cen.ele6))
```