---
title: "Shiny for gene expression clustering"
author: "Michael Hunt"
date: "April 17, 2016"
output: html_document
---

Setup of the expression data and app

(1 point possible)
Let's set up the tissue gene expression data as in the previous video/chapter:

```{r}
library(Biobase)
library(hgu133a.db)
library(ph525x)

esHclust = function(es) {
  emat = t(exprs(es))
  rownames(emat) = sampleNames(es)
  dd = data.frame(emat)
  dfHclust(dd)
}

library(tissuesGeneExpression)
data(tissuesGeneExpression)
tgeES = ExpressionSet(e)
annotation(tgeES) = "hgu133a.db"
pData(tgeES) = tab
featureNames(tgeES) =
  make.names(mapIds(hgu133a.db, keys=featureNames(tgeES),
    keytype="PROBEID", column="SYMBOL"), unique=TRUE)
sampleNames(tgeES) = make.names(tgeES$Tissue, unique=TRUE)
```
How many of the feature names of tgeES are missing (NA)?

```{r}
sum(is.na(featureNames(tgeES)))
```
The originating data are limited to probes with gene identifiers. The make.names call ensures that duplicates get unique suffixes of the form .[n].

### Verifying initial configuration

Run 

```{r}
esHclust(tgeES[1:50,1:50])
```

Set the cutoff height to 2.

What is the average silhouette value for this clustering/partition? (Look at the silh tab after changing the cutoff value).

Answer:  
0.43

```{r}
ex = exprs(tgeES[1:50,1:50])
h = hclust(dist(t(ex[1:2,])), method="ward.D")
cc = cutree(h, h=2)
summary(silhouette(cc, dist(t(ex[1:2,]))))
```

### The 50-gene signature for tissue differentiation

Instead of using the first 50 genes, exploit the ExpressionSet interface to make a targeted investigation. We used limma to identify 50 genes for which a null hypothesis of common mean expression across tissues is rejected. The gene set is identified via

```{r}
library(limma)
mm = model.matrix(~Tissue, data=pData(tgeES))
f1 = lmFit(tgeES, mm)
ef1 = eBayes(f1)
sig50 = rownames(ef1$coef[order(ef1$F,decreasing=TRUE)[1:50],])
```

Now we can run

```{r}
esHclust( tgeES[sig50,] )
```
Set the height for cut to 2. 
The average silhouette for the top 2 genes, Euclidean distance, and ward.D clustering is 0.35.

How many clusters are formed by cutting the tree at 2?

Answer:  
7

### Genes ranked 49th and 50th

Scroll to the bottom of the feature list and check the bottom two genes: EEF1G.2 and KLRC4. Then uncheck the top two genes RPL37A and RPL41. Change the clustering method to ward.D2

What is the average silhouette value for the clustering based on these choices?

Answer:  
0.55  
```{r}
ex = exprs(tgeES[sig50,])
h = hclust(dist(t(ex[49:50,])), method="ward.D2")
cc = cutree(h, h=2)
summary(silhouette(cc, dist(t(ex[49:50,]))))
```  

### Random forests for the 5 gene signature

In the book chapter we showed how the random forests algorithm could be used to assess the set of 50 genes with greatest F values for discriminating tissue-specific means. Try it with the 5 gene signature:

```{r}
sig5 = c("IL26", "ZNF674", "UBC.1", "C7orf25.1", "RPS13")
library(MLInterfaces)
library(randomForest)
set.seed(1234)
rf2 = MLearn(Tissue~., tgeES[sig5,], randomForestI, xvalSpec("NOTEST"))
RObject(rf2)
```
In percent, what is the out of bag estimate of the misclassification error rate?