---
title: "Reference Genomes"
author: "Michael Hunt"
date: "April 10, 2016"
output: html_document
---

### Reference genome discovery

How many Bioconductor packages provide reference genomic sequence for `zebrafish` (_Danio rerio_)? Exclude the packages with `suffix .masked`, that we will discuss later.:


```{r}
library(Biostrings)
available.genomes()
grep("mask", grep("Drerio", available.genomes(), value=TRUE), invert=TRUE, value=TRUE) # exclude masked
```

### Masking structures for genome gaps and repetitions

We have noted that the reference genome builds for complex organisms are works in progress. Genomic sequence "mask" structures have been defined to isolate ambiguous, unmappable, and low-complexity segments of genomes so that sequence analysis research can be targeted to reflect current knowledge of sequence regions that are more likely to be functionally informative.

Obtain BSgenome.Hsapiens.UCSC.hg19.masked (it is only a 20MB transfer.)

Run the commands

```{r}
library(BSgenome.Hsapiens.UCSC.hg19.masked)
c17m = BSgenome.Hsapiens.UCSC.hg19.masked$chr17
```
What is the class of c17m?

```{r}
class(c17m)
```

### Quantifying assembly gaps


When we print out the value of a MaskedDNAString we get a report on types of mask present. Part of the report for chromosome 17 in hg19 is:

`
    81195210-letter "MaskedDNAString" instance (# for masking)  
seq: AAGCTTCTCACCCTGTTCCTGCATAGATAATTGCAT...GTGGGTGTGGGTGTGGTGTGTGGGTGTGGGTGTGGT  
masks:  
  maskedwidth maskedratio active names                               desc  
1     3400000  0.04187439   TRUE AGAPS                      assembly gaps  
...  
`
In build hg19, what percentage of the length of chromosome 22 is occupied by "assembly gaps"? Reply with an integer between 0 and 100.

```{r chr 22}
c22m = BSgenome.Hsapiens.UCSC.hg19.masked$chr22
round(100*sum(width(masks(c22m)$AGAPS))/length(c22m),0)
```
